---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, message=FALSE, echo=FALSE}
library(dplyr)
library(lubridate)
library(lattice)

files <- list.files()
if (!("activity.csv" %in% files)){
        unzip("activity.zip")
        print("hola")
}
rm(files)
knitr::opts_chunk$set(message = FALSE)
```
```{r}
data <- read.csv("activity.csv");
head(data)
```

## What is mean total number of steps taken per day?
```{r, message=FALSE}
steps_by_day <- data %>% group_by(date) %>% 
        summarise(date = unique(date),total_steps = sum(steps,na.rm = T))
hist(steps_by_day$total_steps)
```

Mean of the steps by day
```{r}
mean(steps_by_day$total_steps)
```

Median
```{r}
median(steps_by_day$total_steps)
```

## What is the average daily activity pattern?
```{r, message=FALSE}
avg_interval <- data %>% group_by(interval) %>% 
        summarise(interval = unique(interval),mean_steps = mean(steps,na.rm = T))
plot(avg_interval, type = 'l')
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
avg_interval[which.max(avg_interval$mean_steps),]$interval
```
is the interval number **835**

## Imputing missing values
```{r}
table(is.na(data$steps))[[2]]
```
There are **2304** missing values


The strategy is input the mean for that 5-minute interval

```{r}
datafull <- data
for(i in 1:nrow(data)){
        if (is.na(datafull[i,]$steps)){
                datafull[i,]$steps <- avg_interval[i,]$mean_steps
        }
}
head(datafull)
```
```{r, message=FALSE}
steps_by_day <- datafull %>% group_by(date) %>% 
        summarise(date = unique(date),total_steps = sum(steps,na.rm = T))
hist(steps_by_day$total_steps)
```

Mean of the steps by day
```{r}
mean(steps_by_day$total_steps)
```

Median
```{r}
median(steps_by_day$total_steps)
```

*Do these values differ from the estimates from the first part of the assignment? 

Yes, there is an increase in the mean and median, in addition to some intervals of the histogram

## Are there differences in activity patterns between weekdays and weekends?
```{r}
datafull$date <- as.Date(datafull$date)
datafull <- mutate(datafull,week_time = ifelse(wday(date) %in% c(1,2),"weekend","weekday"))
week <- split(datafull, datafull$week_time)
weekend <- week[[2]]
weekday <- week[[1]]
```

Weekend: 
```{r}
weekend_interval <- weekend %>% group_by(interval) %>% 
        summarise(interval = unique(interval),mean_steps = mean(steps,na.rm = T))
plot(weekend_interval, type = 'l', main = "Weekend")
```


Weekday: 
```{r}
weekday_interval <- weekday %>% group_by(interval) %>% 
        summarise(interval = unique(interval),mean_steps = mean(steps,na.rm = T))
plot(weekday_interval, type = 'l', main = "Weekday")
```
